---
title: "ggplot2: elegant graphics for data analysis"
output: html_notebook
---

##purpose of this section

* Provide overview of plotting frameworks in R
* Give an introduction to ggplot2 library
* Learn how to produce various graphics with ggplot2

##software

* R version 3.4.0
* R Studio 1.0.143
* base, graphics, lattice, ggplot2, ggthemes, ggalt, ggrepel, raster

##what libraries we can use to visualize data in R?

* base plotting system
* lattice
* ggplot2
* ggvis, vegalite, plotly

##base plotting system: simple scatterplot

```{r}
View(cars)
plot(cars)
```

##base plotting system: line graph

```{r}
plot(EuStockMarkets)
```

##base plotting system: scatterplot matrix

```{r}
plot(iris)
```

##base plotting system: customisation

```{r}
plot(cars, xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     las = 1)
lines(lowess(cars$speed, cars$dist), col = "red")
title(main = "cars data")
```

##base plotting system: pros and cons

###pros

* short and simple syntax 
* generic functions
* good for quick and dirty dataviz

###cons

* hard to customise
* hard to get good looking graphics
* hard to get outside-the-box plots
* no active development and new features

##lattice

```{r}
library(lattice)
```

##lattice: scatterplot

```{r}
xyplot(cars$dist ~ cars$speed)
```

##lattice: line graph

```{r}
xyplot(EuStockMarkets)
```



##lattice: scatterplot matrix

```{r}
splom(iris)
```

##lattice: pros and cons

###pros

* good for statistical graphics
* more flexible than base plotting system

###cons

* no major developments
* still hard to produce outside-the-box graphics

##ggplot2

```{r}
library(ggplot2)
```

##ggplot2: scatterplot

```{r}
ggplot(data = cars, aes(x = dist, y = speed))+
  geom_point()
```

##ggplot2: pros and cons

###pros

* based on Grammar of Graphics
* provides good defaults
* provides controls for almost every aspect of the plot
* active development

###cons

* complex syntax
* high entry treshold

##ggplot2: plot structure

```{r eval=FALSE}
ggplot(data, aes(x = x, y = y, color = color, size = size), alpha = '')+
  geom_()+
  scale_x_()+
  scale_y_()+
  scale_color_()+
  scale_size_()+
  coord_()+
  facet_()+
  labs(title = '', subtitle = '', caption = '')+
  theme_minimal(base_family = '', base_size = '')+
  theme(
    legend.position = '',
    plot.background = element_rect(fill = ''),
    plot.title = element_text(),
    plot.subtitle = element_text(),
    plot.caption = element_text(),
    plot.margin = unit()
  )

```

##ggplot2: plot structure

```{r}
g <- ggplot(cars, aes(x = speed, y = dist))+
  geom_point()

str(g)
```

##ggplot2: data

**Data - data frame with variables to be plotted**

```{r}
ggplot(iris)
```


##ggplot2: aestethics

**Aesthetics - scales on which data is mapped**

```{r}
ggplot(iris, aes(x = Sepal.Length))

```

##ggplot2: available aesthetics

```{r}
ggplot2:::.all_aesthetics
```

##ggplot2: geometry

**Geometry - shapes used to represent your data**

```{r}
ggplot(iris, aes(x = Sepal.Length))+
  geom_histogram()
```

##ggplot2: encoding data

```{r}
ggplot(iris, aes(x = Sepal.Length, fill = Species))+
  geom_histogram()
```

##ggplot2: decorating data

```{r}
ggplot(iris, aes(x = Sepal.Length))+
  geom_histogram(fill = 'darkcyan')
```

##ggplot2: available geometries

```{r}
ls(pattern = '^geom_', env = as.environment('package:ggplot2'))
```

##ggplot2: geometries

```{r}
ggplot(presidential)+
  geom_linerange(aes(x = reorder(name, start), ymin = start, ymax = end, color = party), size = 2)
```
View(presidential)


##ggplot2: geometries

```{r}
ggplot(economics, aes(date, uempmed))+
  geom_line()
```

##ggplot2: position

```{r}
ggplot(iris, aes(x = Sepal.Length, fill = Species))+
  geom_histogram(position = 'stack')
```

```{r}
ggplot(iris, aes(x = Sepal.Length, fill = Species))+
  geom_histogram(position = 'dodge')
```

```{r}
ggplot(iris, aes(x = Sepal.Length, fill = Species))+
  geom_histogram(position = 'fill')
```

##ggplot2: positions

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species))+
  geom_point(position = 'jitter')
```

##ggplot2: available positions

```{r}
ls(pattern = '^position_', env = as.environment('package:ggplot2'))
```

##ggplot2: combining geometries

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species))+
  geom_point()+
  geom_smooth()
```

##ggplot2: combining geometries

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Width, color = Species))+
  geom_point(size = 2)+
  geom_smooth(method = 'lm', se = FALSE, size = 0.5)
```

##ggplot2: combining geometries

```{r}
ggplot(economics, aes(date, psavert))+
  geom_area(fill = 'darkcyan', alpha = 0.2)+
  geom_line(color = 'darkcyan', size = 0.5)
```

##ggplot: combining geometries and data sources

```{r}
ukraine <- raster::getData(name = "GADM", country = "UKR", level = 1)
ukraine <- fortify(ukraine)

cities <- read.csv("https://raw.githubusercontent.com/andriy-gazin/population-map/master/geocoded.csv", 
stringsAsFactors = F)

ggplot(ukraine)+
  geom_polygon(aes(long, lat, group = group), 
               color = "#3A3F4A", fill = "#EFF2F4", size = 0.1)+
  geom_point(data = cities,
             aes(lon, lat), alpha = 0.75, size = 0.5)
```


##ggplot2: statistics

**Statistics is a way of transforming data during plotting**

```{r eval=F}
ggplot(iris, aes(x = Sepal.Length, fill = Species))+
  geom_bar(stat = 'count')
```

##ggplot2: statistics

```{r eval=F}
ggplot(mtcars)+
  geom_bar(aes(x = row.names(mtcars), y = mpg), stat = 'identity')
```

##ggplot2: statistics

```{r}
ggplot(diamonds, aes(x = carat, y = price))+
  geom_point()
```

```{r}
ggplot(diamonds, aes(x = carat, y = price))+
  geom_point()+
  stat_binhex()
```

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Petal.Width))+
  geom_point()+
  stat_sum()
```

##ggplot2: available stats

```{r}
ls(pattern = '^stat_', env = as.environment('package:ggplot2'))
```

##ggplot2: scales

**Scale is a way of transforming data points to screen points**

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Petal.Width))+
  geom_count(aes(color = ..n..))+
  scale_size_area(max_size = 5) +
  viridis::scale_color_viridis(option="magma")
```

##ggplot2: scales

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Petal.Width, color = Species))+
  geom_point()+
  scale_color_brewer(type = 'qual', palette = 2)
```
##ggplot2: scales

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Petal.Width, color = Species))+
  geom_point()+
  scale_color_manual(values = c('red', 'green', 'blue'))
```

##ggplot2: scales

```{r}
ggplot(diamonds, aes(x = carat, y = price))+
  geom_point(alpha = 0.1)+
  scale_y_log10()
```

##ggplot2: scales

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Petal.Width, color = Species))+
  geom_point()+
  scale_x_continuous(minor_breaks = NULL)+
  scale_y_reverse(breaks = seq(0,2.5,0.25), labels=paste(seq(0,2.5,0.25)))+
  scale_color_manual(values = c('red', 'green', 'blue'))
```

##ggplot2: available scales

```{r}
ls(pattern = '^scale_', env = as.environment('package:ggplot2'))
```


##ggplot2: coordinates

**Coordinates is a way of positioning screen points**

```{r}
ggplot(mtcars)+
  geom_bar(aes(x = row.names(mtcars), y = mpg), stat = 'identity')
```

##ggplot2: coordinates, coord_flip

```{r}
ggplot(mtcars)+
  geom_bar(aes(x = reorder(row.names(mtcars), mpg), y = mpg), stat = 'identity')+
  coord_flip() + labs(x="Car Models", y="MPG", caption="Source of the Data", title="MPG for each Car Model")
```

##ggplot2: coordinates

```{r}
ggplot(presidential)+
  geom_linerange(aes(x = reorder(name, start), ymin = start, ymax = end, color = party), size = 2)+
  coord_flip()+
  
```


##ggplot2: available coordinates

```{r}
ls(pattern = '^coord_', env = as.environment('package:ggplot2'))
```


##ggplot2: facets

**Facets are rows and columns of subplot **

```{r}
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)
```

##ggplot2: facets

```{r}
ggplot(diamonds)+
  geom_point(aes(x = carat, y = price))+
  facet_wrap(~color, nrow=1)
```

##ggplot2: theme

**Theme is a way of decorating plot**

```{r}
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)+
  theme_gray()
```

##ggplot2: built-in themes

```{r}
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)+
  theme_dark()
```

```{r}
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)+
  theme_linedraw()
```

##ggplot2: available themes

```{r}
ls(pattern = '^theme_', env = as.environment('package:ggplot2'))
```

##ggplot2: extra ggthemes package

```{r}
library(ggthemes)
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)+
  theme_fivethirtyeight()
```

```{r}
library(ggthemes)
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)+
  theme_wsj()
```

```{r}
library(ggthemes)
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width))+
  facet_wrap(~Species)+
  theme_economist()
```

```{r}
ukraine <- raster::getData(name = "GADM", country = "UKR", level = 1)
ukraine <- fortify(ukraine)

cities <- read.csv("https://raw.githubusercontent.com/andriy-gazin/population-map/master/geocoded.csv", 
stringsAsFactors = F)

ggplot(ukraine)+
  geom_polygon(aes(long, lat, group = group), 
               color = "#3A3F4A", fill = "#EFF2F4", size = 0.1)+
  geom_point(data = cities,
             aes(lon, lat), alpha = 0.75, size = 0.5)+
  theme_map()
```


##ggthemes: available themes

```{r}
library(ggthemes)
ls(pattern = '^theme_', env = as.environment('package:ggthemes'))
```

##ggplot2: custom theme

```{r}
library(ggthemes)
ggplot(iris)+
  geom_point(aes(x = Sepal.Length, y = Sepal.Width), color = 'darkcyan', size = 2)+
  facet_wrap(~Species)+
  theme_minimal()+
  theme(
    text = element_text(family = 'Ubuntu Condensed', size = 12),
    rect = element_rect(fill = '#ece2f0'),
    line = element_line(linetype = 'dotted', size = 0.2),
    strip.text = element_text(face = 'bold', size = rel(1.1)),
    panel.spacing.x = unit(1.1, 'lines'),
    axis.text = element_text(size = rel(1.01)),
    axis.title = element_blank(),
    panel.grid.major = element_line(color = '#000000'),
    plot.margin = unit(c(1, 1, 1, 1), 'cm'),
    plot.background = element_rect()
  )
```

##ggplot2: extensions

* ggalt
* ggraph
* ggmap
* geofacet
* viridis
* scales
* ggrepel
* gganimate
* ggmosaic
* gganimate
* ggjoy
* ggthemes
* ggthemer
* ggedit

##ggalt
```{r}
library(ggalt)
ggplot(mtcars)+
  geom_lollipop(aes(x = reorder(row.names(mtcars), mpg), y = mpg))+
  coord_flip()

```

##ggrepel

```{r}
ggplot(mtcars, aes(x = mpg, y = hp))+
  geom_point()+
  geom_text(aes(label = rownames(mtcars), family = 'Ubuntu Condensed'))
```

```{r}
library(ggrepel)
ggplot(mtcars, aes(x = mpg, y = hp))+
  geom_point()+
  geom_text_repel(aes(label = rownames(mtcars), family = 'Ubuntu Condensed'))
```

##viridis

```{r}
library(viridis)
ggplot(diamonds, aes(x = carat, y = price))+
  geom_hex()+
  scale_fill_viridis()
```

##where to learn about ggplot2

* ggplot2 docs
* stackoverflow