---
title: "Exploratory Data Analysis with ggplot2"
output: html_notebook
---

##read the data

```{r}
df <- read.csv('weather_2016.csv', stringsAsFactors = F)
```

##inspect the data

```{r}
head(df)
```

```{r}
tail(df)
```

```{r}
str(df)
```

##lets reduce number of variables

```{r}
library(dplyr)
df <- df %>% 
  select(date.year:date.min, tempm, hum, pressurem, conds, wdire, wdird, wspdm)
```

##we need proper dates

```{r}
df <- df %>% 
  mutate(date = paste(date.year, date.mon, date.mday, sep='-'),
         time = paste(date.hour, date.min, sep = ':'),
         datetime = paste(date, time, sep = ' '),
         datetime = as.POSIXct(strptime(datetime, format = '%Y-%m-%d %H:%M')),
         month = lubridate::month(datetime, label = TRUE, abbr = FALSE))
```

##get seasons

```{r}
df$season <- case_when(
  df$month %in% c('January', 'February', 'December') ~ 'winter',
  df$month %in% c('March', 'April', 'May') ~ 'spring',
  df$month %in% c('June', 'July', 'August') ~ 'summer',
  df$month %in% c('September', 'October', 'November') ~ 'autumn'
)

df$season <- factor(df$season, ordered = TRUE, 
                    levels = c('winter', 'spring', 'summer', 'autumn'))
```

##ok, let's go exploring

```{r}
library(ggplot2)
ggplot(df, aes(x = tempm, fill=season))+
  geom_histogram(binwidth = 1)
```

##color by season

```{r}
ggplot(df, aes(x = tempm, fill = season))+
  geom_histogram(binwidth = 1)
```

##color by season

```{r}
ggplot(df, aes(x = tempm, fill = season))+
  geom_density(alpha = 0.3)
```

##or lets change representation

```{r}
ggplot(df, aes(x = season, y = tempm, fill=season))+
  geom_jitter(alpha=0.2) +
  geom_boxplot(alpha=0.8)
  
```

##once again

```{r}
ggplot(df, aes(x = season, y = tempm))+
  geom_jitter(alpha = 0.1)+
  geom_boxplot(fill = NA, color = 'white', size = 1)
```

##and again

```{r}
ggplot(df, aes(x = season, y = tempm))+
  geom_violin()
```

##and once more

```{r}
library(ggbeeswarm)
ggplot(df, aes(x = season, y = tempm, color = tempm))+
  geom_quasirandom(alpha = 0.1)+
  viridis::scale_color_viridis()
```

##and once more

```{r}
library(ggbeeswarm)
ggplot(df, aes(x = month, y = tempm, color = tempm))+
  geom_quasirandom(alpha = 0.1)+
  viridis::scale_color_viridis()
```

##show variation during the day

```{r}
ggplot(df, aes(x = as.Date(datetime), y = time, fill = tempm))+
  geom_tile()
```


##no, really, show variation during the day
```{r}
df %>% filter(stringr::str_detect(df$time, '00|30')) %>% 
ggplot(aes(x = as.Date(datetime), 
               y = as.POSIXct(strptime(time, format = '%H:%M')), 
               fill = tempm))+
  geom_tile()+
  scale_y_datetime(date_labels = '%H-%M', date_breaks = '4 hour') +
  viridis::scale_fill_viridis()
```

##can we do it another way? sure we can

```{r}
df %>% filter(stringr::str_detect(df$time, '00|30')) %>% 
ggplot(aes(x = as.POSIXct(strptime(time, format = '%H:%M')), 
           y = tempm))+
  geom_point(alpha=0.1)+geom_smooth()+
  scale_x_datetime(date_labels = '%H:%M')+
  facet_wrap(~month)
```

##summarise by weather conditions

```{r}
df %>% group_by(conds) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = reorder(conds, count), y = count))+
  geom_bar(stat = 'identity')+
  coord_flip()
```

##show only rainy days

```{r}
df %>% filter(stringr::str_detect(conds, pattern = 'Rain')) %>% 
  ggplot(aes(x = datetime))+
  geom_linerange(aes(ymin = -1, ymax = 1), alpha = 0.2)
```

##show only snowy days

```{r}
df %>% filter(stringr::str_detect(conds, pattern = 'Snow')) %>% 
  ggplot(aes(x = datetime))+
  geom_linerange(aes(ymin = -1, ymax = 1, color = tempm), alpha = 0.2)+
  viridis::scale_color_viridis()
```

##show 'clear' vs 'cloudy'

```{r}
df %>% filter(stringr::str_detect(conds, pattern = 'Clear|Cloud')) %>% 
  ggplot(aes(x = datetime))+
  geom_linerange(aes(ymin = -1, ymax = 1, color = conds))+
  scale_color_brewer(type = 'qual')
```

##show 'clear' vs 'cloudy'

```{r}
df %>% filter(stringr::str_detect(conds, pattern = 'Clear|Cloud')) %>% 
  ggplot(aes(x = datetime))+
  geom_linerange(aes(ymin = -1, ymax = 1, color = conds), size = 0.1)+
  scale_color_brewer(type = 'qual', palette = 2)+
  facet_wrap(~conds, ncol = 1)
```

##explore winds data

```{r}
df %>% group_by(wdire) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(reorder(wdire, count), count))+
  geom_bar(stat = 'identity')+
  coord_flip()
```

##explore winds data

```{r}
df %>% group_by(wdire, season) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(reorder(wdire, count), count))+
  geom_bar(stat = 'identity')+
  coord_flip()+
  facet_wrap(~season, ncol = 4)
```

##wind direction by season

```{r}
ggplot(df, aes(x = wdird))+
  geom_histogram()+
  facet_wrap(~season) +
  coord_polar()
```

##wind speed by season

```{r}
ggplot(df, aes(x = wspdm))+
  geom_histogram()+
  facet_wrap(~season)
```

##wind direction vs wind speed

```{r}
ggplot(df, aes(x = wdird, y = wspdm))+
  geom_point(alpha = 0.25)+
  facet_wrap(~season)
```

##wind direction vs wind speed

```{r}
ggplot(df, aes(x = wdird, y = wspdm))+
  geom_count(alpha = 0.5)+
  facet_wrap(~season)
```

##wind direction vs wind speed

```{r}
ggplot(df, aes(x = wdird, y = wspdm))+
  geom_count(aes(color = ..n..), alpha = 1)+
  viridis::scale_color_viridis(option = 'magma', direction = -1)+
  facet_wrap(~season)
```

##wind direction vs wind speed

```{r}
ggplot(df, aes(x = wdire, y = wspdm))+
  geom_count(aes(color = ..n..), alpha = 1)+
  viridis::scale_color_viridis(option = 'magma', direction = -1)+
  coord_flip()+
  facet_wrap(~season)
```

##wind speed by month

```{r}
library(ggjoy)
ggplot(df, aes(x = wspdm, y = month))+
  geom_joy()
```

##wind speed by month

```{r}
library(forcats)
ggplot(df, aes(x = wspdm, y = fct_rev(month)))+
  geom_joy()
```

##daily means
```{r}
df %>% mutate(date = as.Date(datetime)) %>% 
  group_by(date) %>% 
  summarise(temp_avg = mean(tempm, na.rm = T)) %>% 
  ggplot(aes(x = date, y = temp_avg))+
  geom_line()
```

##daily ranges
```{r}
df %>% mutate(date = as.Date(datetime)) %>% 
  group_by(date) %>% 
  summarise(temp_min = min(tempm, na.rm = T),
            temp_max = max(tempm, na.rm = T),
            temp_avg = mean(tempm, na.rm = T)) %>% 
  ggplot(aes(x = date))+
  geom_linerange(aes(ymin = temp_min, ymax = temp_max))
```


##daily ranges and daily means
```{r}
df %>% mutate(date = as.Date(datetime)) %>% 
  group_by(date) %>% 
  summarise(temp_min = min(tempm, na.rm = T),
            temp_max = max(tempm, na.rm = T),
            temp_avg = mean(tempm, na.rm = T)) %>% 
  ggplot(aes(x = date))+
  geom_linerange(aes(ymin = temp_min, ymax = temp_max), 
                 color = 'gray', alpha = 0.7)+
  geom_line(aes(y = temp_avg), color = 'darkcyan', size = 0.5)
```

##daily ranges and daily means
```{r}
df %>% mutate(date = as.Date(datetime)) %>% 
  group_by(date) %>% 
  summarise(temp_min = min(tempm, na.rm = T),
            temp_max = max(tempm, na.rm = T),
            temp_avg = mean(tempm, na.rm = T)) %>% 
  ggplot(aes(x = date))+
  geom_ribbon(aes(ymin = temp_min, ymax = temp_max), 
                 fill = 'gray', alpha = 0.7)+
  geom_line(aes(y = temp_avg), color = 'darkcyan', size = 0.3)
```

##temp vs humidity

```{r}
ggplot(df, aes(hum, tempm))+
  geom_jitter()
```

##temp vs humidity

```{r}
ggplot(df, aes(pressurem, tempm))+
  geom_jitter()
```

##customize

```{r}

```